Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity\vspace{-7ex}

### 2016-08-06 papers/kmeans/response_to_referee.txt

Because of discussions and checking of results that happened
subsequent to our submission of this paper with the APOGEE team, some
of which made use of proprietary data and information about the
survey, we have expanded the author list in accordance with the
APOGEE/SDSS-III authorship policies.  The consequence is that there
are now several additional new authors, most of whom contributed to
this project most directly by building the APOGEE project (and hence
the input data and tests thereof).

We added the two crucial APOGEE references: Eisenstein 2011 (SDSS-III)
Holtzman 2015 (APOGEE DR12).

We softened very slightly the "no previous successes" sentence in the
abstract and also cited Majewski et al 2012. We also added a caveat
that globular-cluster members are over-represented in APOGEE because
of targeting decisions.

We fixed a small bug with the cluster density measure and updated the
figure. This changes no conclusions.

We fixed the (embarrassing) bug that everywhere we say "Pal 5" we mean
"M5". This was a straight-up mistake in the paper, and leads to some
changes in the discussion, including some reference changes. No
fundamental conclusions change because of this.

### 2016-08-06 papers/kmeans/kmeans.tex
requested citations

This idea---dubbed ``chemical tagging''
\citep{freeman}---is one of the principal motivations for a number
of surveys, including \apogee\ \citep{apogee},
\gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
In order to determine the precise abundance labels for chemical
tagging, these surveys are each measuring high-resolution, high
signal-to-noise ($S/N$) spectra for hundreds of thousands of stars across the
Galaxy's disk, bulge and halo.

@@

There are, however, two big issues.
The first is that the physical assumptions behind the idea may require
refinement:
There may be chemical-abundance overlaps among open clusters
\citep{blancocuaresma}, coeval groups of stars may have similar tags
but different birth places \citep{mitschang}, and the
chemical-abundance space might be low in dimensionality.
On the other hand, precise studies of stellar twins \citep{melendez, jofre}
and open clusters \citep{bovy} indicate that chemical abundances ought to be
very similar for co-eval stars, and peculiar abundance ratios
have been successfully used to identify disrupted cluster members (for example, \citealt{ocen}).

@@

We chose a few interesting cases from the high-density clusters
in the $K=256$ run and show them in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
\ref{fig:halo}, and \ref{fig:disk}.
The first three of these are dominated by stars in the known clusters
M13, Pal5, and Sagittarius.
The fourth is a halo structure with high velocity dispersion and
possibly accreted.
The fifth is a thin-disk star-formation feature.
We will discuss the astrophysical implications of these results in
\sectionname~\ref{sec:discussion}.

@@

\begin{figure}[!bp]
\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster (which is displayed in more detail in 
  \figurename~\ref{fig:M13}) is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense $K=256$
  abundance-space cluster have been rendered as unique, prominent
  triangles (color is velocity rank, orientation is $\logg$ rank).
  The lower-right plot shows this cluster (circle) in context of the
  other clusters (dots).  This cluster, which was identified only in
  abundance space (six projections of which are the top six panels in
  this \figurename), turns out to be dominated by the halo globular
  cluster M13. The symbol orientations and colors have no meaning;
  they are randomly generated---one unique color and orientation for
  each highlighted star---to make the points cross-identifiable across
  panels.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius dwarf spheroidal galaxy.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a hitherto unrecognized high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk, probably one that has been discovered previously \citep{wegg}.\label{fig:disk}}
\end{figure}

@@

There are scant abundance measurements reported for Palomar 5, for which
most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
metallicity of [Fe/H]$ = -1.3$, which lies between the only
photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
Indeed, this work contains the largest number of abundances for Palomar 5,
permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
that have retained the unique cluster abundance signature.

@@

The \apogee\ project shows great promise for these studies, and with
the appearance of the companion papers, we will release the
chemical-abundance measurements we used here for further study.
The future is even brighter, however: \apogee\ is expanding to more measured
elements, more stars, and all-sky angular coverage with \apogee2, and
\galah\ is working towards releasing chemical abundances on a larger set
of 29 elements.
The \apogee\ elements employed in this project include alpha
(Mg, O, Si, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
and iron peak (Fe, Ni, V, Mn) elements.
\galah\ will deliver element abundances from all the major
nucleosynthetic processes, including light proton capture elements,
alpha, odd-Z, iron-peak, light and heavy slow neutron capture, and
rapid neutron capture.
Chemical tagging capabilities are expected to grow as the
\emph{product} of the measured nucleosynthetic pathways.

\acknowledgements
It is a pleasure to thank
  Katia Cunha (NOAO),
  Jeremy Magland (\acronym{SCDA}),
  Don Schneider (PSU),
  Keivan Stassun (Vanderbilt),
  and the Blanton--Hogg group meeting
for valuable comments and discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1312863}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

@@

@@

>>>>

This idea---dubbed ``chemical tagging''
\citep{freeman, jbh}---is one of the principal motivations for a number
of surveys, including \apogee\ \citep{apogee},
\gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
In order to determine the precise abundance labels for chemical
tagging, these surveys are each measuring high-resolution, high
signal-to-noise ($S/N$) spectra for hundreds of thousands of stars across the
Galaxy's disk, bulge and halo.

@@

There are, however, two big issues.
The first is that the physical assumptions behind the idea may require
refinement:
There may be chemical-abundance overlaps among open clusters
\citep{blancocuaresma}, coeval groups of stars may have similar tags
but different birth places \citep{mitschang}, and the
chemical-abundance space might be low in dimensionality.
On the other hand, precise studies of stellar twins \citep{melendez, jofre}
indicate that pairs of stars can be found with unusually similar abundances, 
open clusters show remarkably uniform chemical abundances \citep{bovy},
and peculiar abundance ratios
have been successfully used to identify disrupted cluster members (for example, \citealt{ocen}).
There have also been hints seen of relationships between chemistry and kinematics
(for example, \citealt{helmi, helmi2}).

@@

We chose a few interesting cases from the high-density clusters
in the $K=256$ run and show them in
\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
\ref{fig:halo}, and \ref{fig:disk}.
The first three of these are dominated by stars in the known clusters
M13 and M5, and the Sagittarius stream (we identify overlap with these objects
by looking at stellar position and velocity, and, in some cases,
\apogee\ targeting flags).
The fourth is a halo structure with high velocity dispersion and
possibly accreted.
The fifth is a thin-disk star-formation feature.
We will discuss the astrophysical implications of these results in
\sectionname~\ref{sec:discussion}.

@@

\begin{figure}[!bp]
\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
\caption{The distribution of membership and density for the 256
  abundance-space clusters returned by the k-means algorithm at
  $K=256$.  The densest cluster (which is displayed in more detail in 
  \figurename~\ref{fig:M13}) is far more dense than the second
  densest.\label{fig:densities}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of the most dense $K=256$
  abundance-space cluster have been rendered as unique, prominent
  triangles (color is velocity rank, orientation is $\logg$ rank).
  The lower-right plot shows this cluster (circle) in context of the
  other clusters (dots).  This cluster, which was identified only in
  abundance space (six projections of which are the top six panels in
  this \figurename), turns out to be dominated by the halo globular
  cluster M13. The symbol orientations and colors have no meaning;
  they are randomly generated---one unique color and orientation for
  each highlighted star---to make the points cross-identifiable across
  panels.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by globular
  cluster M5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius dwarf spheroidal galaxy.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a hitherto unrecognized high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk, probably one that has been discovered previously \citep{wegg}.\label{fig:disk}}
\end{figure}

@@

%HOGG: THIS PARAGRAPH IS ALL WRONG!
There are scant abundance measurements reported for Palomar 5, for which
most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
metallicity of [Fe/H]$ = -1.3$, which lies between the only
photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
Indeed, this work contains the largest number of abundances for Palomar 5,
permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
that have retained the unique cluster abundance signature.

@@

The \apogee\ project shows great promise for these studies, and with
the appearance of the companion papers, we will release the
chemical-abundance measurements we used here for further study.
The future is even brighter, however: \apogee\ is expanding to more measured
elements, more stars, and all-sky angular coverage with \apogee2, and
\galah\ is working towards releasing chemical abundances on a larger set
of 29 elements.
The \apogee\ elements employed in this project include alpha
(Mg, O, Si, S, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
and iron peak (Fe, Ni, V, Mn) elements.
\galah\ will deliver element abundances from all the major
nucleosynthetic processes, including light proton capture elements,
alpha, odd-Z, iron-peak, light and heavy slow neutron capture, and
rapid neutron capture.
Chemical tagging capabilities are expected to grow as the
\emph{product} of the measured nucleosynthetic pathways.

\acknowledgements
It is a pleasure to thank
  Joss Bland-Hawthorn (Sydney),
  Jo Bovy (Toronto),
  Charlie Conroy (Harvard),
  Katia Cunha (\acronym{NOAO}),
  Amina Helmi (Kapteyn),
  Jeremy Magland (\acronym{SCDA}),
  Don Schneider (\acronym{PSU}),
  Keivan Stassun (Vanderbilt),
  Angus Williams (Cambridge),
  and the Blanton--Hogg group meeting
for valuable comments and discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1312863}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

@@

\bibitem[Bland-Hawthorn \etal(2010)]{jbh}
Bland-Hawthorn,~J., Krumholz,~M.~R., \& Freeman,~K., 2010, \apj, 713, 166

@@

\bibitem[Helmi \etal(2006)]{helmi}
Helmi,~A., Navarro,~J.~F., Nordstr{\"o}m,~B., Holmberg,~J., Abadi,~M.~G., \& Steinmetz,~M., 2006, \mnras, 365, 1309
\bibitem[Helmi \etal(2014)]{helmi2}
Helmi,~A., Williams,~M., Freeman,~K.~C., Bland-Hawthorn,~J., \& De~Silva,~G., 2014, \apj, 791, 135
